Library preparation.
```{r,warning=FALSE,message=FALSE}
library(reticulate)
library(stringr)
library(dplyr)
library(tidyverse)
library(devtools)
library(ggbiplot)
library(glmnet)
library(RColorBrewer)
library(cowplot)
```

First we need to read pickle files into R.
```{r,warning=FALSE}
source_python("R_read_pickle.py")
setwd("/n/holyscratch01/stephenson_lab/Users/taylerli/SCENIC/data")
dta <- read_pickle_file("BM_AML_normal_output/BM_AML_normal_aucell_values.pkl.gz")
```

Then, get the correct annotations of normal cells and place them alongside AUCell data.
```{r}
filenames = list.files(path = "/n/holyscratch01/stephenson_lab/Users/taylerli/SCENIC/data",
                       full.names = T)
filenames1 = filenames[str_detect(filenames, pattern = ".*GSM\\d*_BM.*\\.anno\\.txt\\.gz")]
filenames1 = c(filenames1,
               filenames[str_detect(filenames, pattern = ".*GSM\\d*_AML.*\\.anno\\.txt\\.gz")])
filenames1 <- filenames1[c(1,2,3,4,5,6,7,8,9,10,12,13,14,11,15)]

anno <- data.frame()
for (i in 1:length(filenames1)) {
    anno <- rbind(anno,
                  read.csv(filenames1[i], header = T, row.names = 1, sep = '\t')[, 1:27])
}
anno <- anno[anno$CellType != "",] %>%
    dplyr::select(CellType) %>%
    mutate(HC = ifelse(str_sub(CellType, start = -4) == 'like', 'Cancer', 'Healthy')) %>%
    mutate(HC = relevel(factor(HC), ref = 'Healthy'))

dta <- cbind(anno, dta)
```

LASSO.
```{r}
dta_lasso <- dta[, -1] %>%
    mutate(HC = as.numeric(HC) - 1)
x <- model.matrix(HC ~ ., data = dta_lasso)
y <- dta_lasso$HC

set.seed(292876)
lambda_grid <- .5 ^ (-20:20)
lasso.mod = glmnet(x, y, alpha=1, family="binomial", lambda = lambda_grid)
#par(mfrow=c(1,2))
#plot(lasso.mod)
cv.out <- cv.glmnet(x,y, alpha=1)
#plot(cv.out)

tab <- as.matrix(coef(cv.out, s="lambda.min"))
tab <- tab[which(rowSums(tab)!=0), ][-1]
col_names <- gsub('`', '', names(tab))

t5 <- gsub('`', '', head(names(sort(tab, decreasing = T)), 5))
l5 <- gsub('`', '', tail(names(sort(tab, decreasing = T)), 5))


```

Do a exploratory PCA with color coding for different cell types.
```{r}
colors <- c("HSC" = "black",
            "Prog" = "gray65",
            "GMP" = "darkorchid",
            "ProMono" = "orangered2",
            "Mono" = "gold1",
            "cDC" = "green3",
            "pDC" = "royalblue",
            "earlyEry" = "sienna2",
            "lateEry" = "violetred4",
            "ProB" = "turquoise3",
            "B" = "springgreen2",
            "Plasma" = "seagreen",
            "T" = "skyblue1",
            "CTL" = "maroon3",
            "NK" = "midnightblue"
            )
colors2 <- c("HSC" = "black",
            "Prog" = "gray65",
            "GMP" = "darkorchid",
            "ProMono" = "orangered2",
            "Mono" = "gold1",
            "cDC" = "green3",
            "pDC" = "royalblue",
            "earlyEry" = "sienna2",
            "lateEry" = "violetred4",
            "ProB" = "turquoise3",
            "B" = "springgreen2",
            "Plasma" = "seagreen",
            "T" = "skyblue1",
            "CTL" = "maroon3",
            "NK" = "midnightblue",
            "HSC-like" = "black",
            "Prog-like" = "gray65",
            "GMP-like" = "darkorchid",
            "ProMono-like" = "orangered2",
            "Mono-like" = "gold1",
            "cDC-like" = "green3"
            )
# dta.pca <- prcomp(dta[, 3:ncol(dta)],
#                   center = T, 
#                   scale. = T
#            )
# g <- ggbiplot(dta.pca,
#               obs.scale = 1,
#               var.scale = 1,
#               groups = dta[, c("CellType", "HC")],
#               #ellipse = TRUE,
#               #circle = TRUE,
#               #ellipse.prob = 0.68,
#               var.axes = F,
#               alpha = 0.7
#      )
# g + scale_colour_manual(name="Cell Type", values=colors)
```

Heatmap analysis.
```{r}
dta_heat_1 <- dta
dta_heat_1$CellType <- factor(dta_heat_1$CellType, c("HSC", "Prog", "GMP", "ProMono", "Mono", "cDC",
                                                     "pDC", "earlyEry", "lateEry", "ProB", "B", "Plasma",
                                                     "T", "CTL", "NK", "HSC-like", "Prog-like", "GMP-like",
                                                     "ProMono-like", "Mono-like", "cDC-like"))
dta_heat_1 <- dta_heat_1[order(dta_heat_1$CellType), ]
# col_order <- hclust(dist(t(dta_heat_1[, 2:ncol(dta_heat_1)])))$order
# heatmap(as.matrix(sapply(dta_heat_1[, col_order], as.numeric)),
#         Rowv = NA,
#         Colv = NA,
#         useRaster = T,
#         scale = "n")
heatmap(as.matrix(dta_heat_1[, 3:ncol(dta_heat_1)]), Rowv = NA, useRaster = T, scale = 'none')
```
```{r}
dta_heat_3 <- dta_heat_1[13244: 20148, -2] %>%
    mutate(CellType = factor(CellType, c("HSC-like", "Prog-like", "GMP-like", "ProMono-like", "Mono-like", "cDC-like")))
# select the most variable regulons
var_df = dta_heat_3[, -1] %>% 
    dplyr::summarise(across(.cols = everything(), .fns = var)) %>% 
    pivot_longer(cols = everything()) %>% 
    slice_max(n = 100, order_by = value)
dta_heat_2 <- dta_heat_3 %>% 
    dplyr::select(c("CellType", var_df$name))
mat <- t(as.matrix(dta_heat_2[, -1]))
r_ord <- rev(hclust(dist(mat))$order)

png('heatmap_c.png')
heatmap(mat[r_ord, ],
        Rowv = NA,
        Colv = NA,
        scale = 'none',
        ColSideColors = as.character(`levels<-`(dta_heat_2$CellType, colors5)))
dev.off()

source("heatmap3.R")

png(filename = "heatmap_hc.png")
heatmap.3(mat[r_ord, ],
        Rowv = NA,
        Colv = NA,
        scale = 'none',
        ColSideColors = cbind(as.character(`levels<-`(dta_heat_1$CellType, colors2)),
                              rep("white", 20148),
                              c(rep("yellow", 13243), rep("lightblue", 6905))),
        col=rev(heat.colors(75)),
        key = F)
dev.off()
# cells on columns
# ColSideColors
# anova for each regulon against cell types, keep the ones with a good p-value
# which regulons are the highest in HSCs, which in GMPs, ...
```



Do a tSNE.
```{r}
# dta_tsne <- dta_heat_1[, 2:ncol(dta_heat_1)]
# dta.cor <- cor(dta_tsne)
# 
# dta.cor.hc <- as.dendrogram(hclust(as.dist(1-dta.cor), method = "average"))
# plot(dta.cor.hc)
# 
# heatmap(dta.cor, Rowv = dta.cor.hc, symm=TRUE, zlim=c(-1, 1), col = dta_heat_1$CellType)
shapes = list("Healthy" = "20",
              "Cancer" = "5")

library(Rtsne)
dta_tsne <- dta_heat_1[, 3:ncol(dta_heat_1)]
dta_celltype <- dta_heat_1[, 1]
dta_hc       <- dta_heat_1[, 2]

tsne_results <- Rtsne(dta_tsne, perplexity=30, check_duplicates = FALSE)

tSNE_df <- tsne_results$Y %>% 
  as.data.frame()
colnames(tSNE_df) = c("tSNE1", "tSNE2")
tSNE_df$CellType = dta_celltype
tSNE_df$HC       = dta_hc


colors5 <- c(
            "HSC-like" = "black",
            "Prog-like" = "gray65",
            "GMP-like" = "darkorchid",
            "ProMono-like" = "orangered2",
            "Mono-like" = "gold1",
            "cDC-like" = "green3"
            )
tSNE_df[which(str_detect(rownames(dta_tsne), "BM")), ] %>%
  ggplot(aes(x = tSNE1, 
             y = tSNE2,
             color = CellType))+
  geom_point(size = 0.5)+
  scale_color_manual(name="Cell Type", values=colors)+
  #scale_shape_manual(name="Healthy/Cancer", values=shapes)+
  theme(legend.position="right",
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        axis.title = element_blank())

colors4 <- c("HSC/HSC-like" = "black",
            "Prog/Prog-like" = "gray65",
            "GMP/GMP-like" = "darkorchid",
            "ProMono/ProMono-like" = "orangered2",
            "Mono/Mono-like" = "gold1",
            "cDC/cDC-like" = "green3",
            "pDC" = "royalblue",
            "earlyEry" = "sienna2",
            "lateEry" = "violetred4",
            "ProB" = "turquoise3",
            "B" = "springgreen2",
            "Plasma" = "seagreen",
            "T" = "skyblue1",
            "CTL" = "maroon3",
            "NK" = "midnightblue"
            )
t<-tSNE_df %>%
  mutate(CellType = gsub('-like', '', CellType)) %>%
  mutate(CellType = case_when(CellType == 'HSC' ~ 'HSC/HSC-like',
                              CellType == 'Prog' ~ 'Prog/Prog-like',
                              CellType == 'GMP' ~ 'GMP/GMP-like',
                              CellType == 'ProMono' ~ 'ProMono/ProMono-like',
                              CellType == 'Mono' ~ 'Mono/Mono-like',
                              CellType == 'cDC' ~ 'cDC/cDC-like',
                              T ~ CellType))
tp<-t%>%ggplot(aes(x = tSNE1, 
             y = tSNE2,
             color = CellType,
             shape = HC))+
  geom_point(size = 0.9)+
  scale_color_manual(name="Cell Type", values=colors4)+
  scale_shape_manual(name="Healthy/Cancer", values=list("Healthy" = "20",
                                                        "Cancer" = "2"))+
  theme(legend.position='none',
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        axis.title = element_blank())

color_plot <-  ggplot(data = t) + 
  geom_point(aes(x = tSNE1, y = tSNE2, color = CellType)) +
  theme(legend.position = 'right',
        legend.title = element_blank()) +
  scale_color_manual(name="Cell Type", values=colors4)
color_legend <- cowplot::get_legend(color_plot)

shape_plot <-  ggplot(data = t) + 
  geom_point(aes(x = tSNE1, y = tSNE2, shape = HC)) +
  theme(legend.position = 'top',
        legend.title = element_blank()) +
  scale_shape_manual(name="Healthy/Cancer", values=list("Healthy" = "20",
                                                        "Cancer" = "2"))
shape_legend <- cowplot::get_legend(shape_plot)

cowplot::plot_grid(plotlist = list(tp, color_legend,shape_legend,NULL),
          rel_heights = c(7, 1),
          rel_widths =  c(3.8, 1.2))

colors3 <- list("Healthy" = "yellow", "Cancer" = "red")
tSNE_df %>%
  filter(HC=='Cancer')%>%
  ggplot(aes(x = tSNE1, 
             y = tSNE2,
             color = CellType))+
  geom_point(size = 0.7)+
  scale_color_manual(name="Cell Type", values=colors5)+
  theme(legend.position="right",
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        axis.title = element_blank())
ggsave("tsne_hc.png")
```

Figure out which regulons are specific to a certain cell type.
```{r}
regulon_differential <- tibble(CellType = levels(dta_heat_3$CellType),
                               fir_indicator = NA,
                               sec_indicator = NA,
                               thi_indicator = NA,
                               fou_indicator = NA,
                               fif_indicator = NA,
                               six_indicator = NA,
                               sev_indicator = NA,
                               eig_indicator = NA,
                               nin_indicator = NA,
                               ten_indicator = NA)
regulon_diff_value <- tibble(CellType = levels(dta_heat_3$CellType),
                               fir_indicator = NA,
                               sec_indicator = NA,
                               thi_indicator = NA,
                               fou_indicator = NA,
                               fif_indicator = NA,
                               six_indicator = NA,
                               sev_indicator = NA,
                               eig_indicator = NA,
                               nin_indicator = NA,
                               ten_indicator = NA)

for (ele in levels(dta_heat_3$CellType)) {
    df1 <- dta_heat_3[dta_heat_3$CellType == ele, -1]
    df2 <- dta_heat_3[dta_heat_3$CellType != ele, -1]
    df_diff <- as.data.frame(colMeans(df1) - colMeans(df2))
    colnames(df_diff) <- c("diff")
    df_diff <- df_diff %>%
        mutate(regulon = rownames(df_diff)) %>%
        dplyr::select(c("regulon", "diff"))
    rownames(df_diff) <- 1:dim(df_diff)[1]
    df_diff_ordered <- df_diff[order(desc(abs(df_diff$diff))), ]
    regulon_differential[regulon_differential$CellType == ele, 2:11] <- as.list(df_diff_ordered$regulon[1:10])
    regulon_diff_value[regulon_diff_value$CellType == ele, 2:11] <- as.list(df_diff_ordered$diff[1:10])
}

regulon_differential
```

Color cells based on regulon activity.
```{r}
color_by_regulon <- function(regulon) {
    tSNE_df %>%
    mutate(REGU = dta_heat_1[, regulon]) %>%
    ggplot(aes(x = tSNE1, 
               y = tSNE2,
               color = REGU,))+
    geom_point(size = 0.7)+
    scale_color_gradient(low="white", high="red")+
    theme(legend.position="right",
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          panel.background = element_blank(),
          axis.text.x=element_blank(),
          axis.ticks.x=element_blank(),
          axis.text.y=element_blank(),
          axis.ticks.y=element_blank(),
          axis.title = element_blank())+
    labs(color = regulon)
}

color_by_regulon("CEBPD(+)")
ggsave("tsne_hc_h1.png")
color_by_regulon("YBX1(+)")
ggsave("tsne_hc_h2.png")
color_by_regulon("E2F2(+)")
ggsave("tsne_hc_h3.png")
```

Random Forest analysis.
```{r}
trainset <- read_pickle_file("train_set.pkl.gz")
validationset <- read_pickle_file("validation_set.pkl.gz")
testset <- read_pickle_file("test_set.pkl.gz")
trainset$CellType <- factor(trainset$CellType, c("HSC-like", "Prog-like", "GMP-like", "ProMono-like", "Mono-like", "cDC-like"))
validationset$CellType <- factor(validationset$CellType, c("HSC-like", "Prog-like", "GMP-like", "ProMono-like", "Mono-like", "cDC-like"))
testset$CellType <- factor(testset$CellType, c("HSC-like", "Prog-like", "GMP-like", "ProMono-like", "Mono-like", "cDC-like"))
trainset <- trainset %>% dplyr::select(-c("HC"))
validationset <- validationset %>% dplyr::select(-c("HC"))
testset <- testset %>% dplyr::select(-c("HC"))


library(randomForest)
library(MASS)
library(caret)
library(knitr)
get_accuracy <- function(fit, testset){
  pred <- predict(fit, newdata = testset, type = "class")
  confusionMatrix(table(pred = pred, true = testset$CellType))$overall[1]
}

set.seed(292876)
# in_train = sample(1:nrow(dta_heat_1), size = 0.7*nrow(dta_heat_1))
# trainset = dta_heat_1[in_train, ]
names(trainset) <- make.names(names(trainset))
# testset = dta_heat_1[-in_train, ]
names(validationset) <- make.names(names(validationset))
names(testset) <- make.names(names(testset))

set.seed(292876); fit <- randomForest(CellType ~ ., data = trainset, ntree = 100, mtry = 60); get_accuracy(fit, trainset); get_accuracy(fit, validationset)

variable_importance <- importance(fit) 
tmp <- data_frame(feature = str_replace(rownames(variable_importance), "\\.\\.\\.", "(+)"),
                  Gini = variable_importance[,1]) %>%
                  arrange(desc(Gini))
head(tmp, 10)
```
